Local Spectral Clustering of Density Upper Level Sets

### 2021-09-01 arxiv/examples_section_backup.tex
Cleaned out arxiv folder.

As the following theorem demonstrates, even when Algorithm~\textcolor{red}{1} is reasonably initialized, if the density cluster $\mathcal{C}_{\sigma}^{(1)}$ is sufficiently geometrically ill-conditioned the cluster estimator $\widehat{C}$ will fail to recover $\mathcal{C}_{\sigma}^{(1)}$. Let
\begin{equation}
\label{eqn:lower_set}
\mathcal{L} = \set{(x_1,x_2) \in \mathcal{X}: x_2 < 0}.
\end{equation}
\begin{theorem}
	\label{thm:ppr_lb}
	Suppose $r < \frac{1}{40}\rho \wedge \frac{1}{4}\sigma$,  $\alpha = 65 \Phi_{\Pbb}(\mathcal{L})$, and $(L,U) = (0,1)$ are inputs to Algorithm~\textcolor{red}{1}. Then, for any
	\begin{equation}
	\label{eqn:lb_sample_size}
	n \geq \max\set{\frac{64}{\epsilon^2 \rho \sigma \pi r^2}, \frac{8}{\epsilon}}
	\end{equation}
	the following statement holds: there exists a set $C^{g} \subset \Xbf$ with $\vol(C^g \cap \mathcal{C}_{\sigma}^{(1)}[\Xbf]) \geq \frac{1}{10}\vol_{n,r}(\mathcal{C}_{\sigma}^{(1)}[\Xbf])$ such that for any seed node $v \in C^g$, the estimator $\widehat{C}$ computed by Algorithm~\textcolor{red}{1} has symmetric set difference with $\Csig^{(1)}[\Xbf]$ of volume at least
	\begin{equation}
	\label{eqn:ppr_lb}
	\frac{\sigma \rho}{r^2}\frac{\vol_{n,r}(\widehat{C} \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf])}{n^2} \geq \frac{1}{4} -  c \frac{\sqrt{\frac{\sigma}{\rho}}}{\epsilon^2} \sqrt{ \log\left(\frac{\rho \sigma}{\epsilon^2 r^2}\right)\frac{\sigma}{r}}
	\end{equation}
	with probability at least $1 - c_1 n \exp\set{-c_2n}$, where $c$ is a universal constant and $c_1,c_2$ are constants which do not depend on $n$.
	Consequently, if
	\begin{equation*} 
	\epsilon^2 > \frac{c}{8} \sqrt{\frac{\sigma}{\rho}} \cdot \sqrt{ \log\left(\frac{\rho \sigma}{\epsilon^2 r^2}\right)\frac{\sigma}{r}}
	\end{equation*}
	then with high probability $\frac{\sigma \rho}{r^2}\frac{\vol_{n,r}(\widehat{C} \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf])}{n^2}$ is at least $1/8$. 
\end{theorem}

Note that the factor of $\frac{\sigma}{r^2}$ on the left hand side of \eqref{eqn:ppr_lb} is a constant multiple of $(1/\vol_{\Pbb,r}(\Csig^{(1)}))$; hence the left hand side is directly comparable to $\frac{\vol_{n,r}(\widehat{C} \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf])}{\vol_{n,r}(\Csig^{(1)}[\Xbf])}$, which is the quantity we upper bound in Theorem~\textcolor{red}{1}.

Theorem~\ref{thm:ppr_lb} is stated with respect to a particular hard case, where the density clusters are rectangular subsets of $\Reals^2$. Although we picked this setting to make the theorem statement simple, our results are easily generalized to $\Reals^d$ and to non-rectangular clusters. Additionally, although we state our lower bound with respect to PPR run on neighborhood graph, the conclusion is likely to hold for a much broader class of spectral clustering algorithms. In the proof of Theorem~\ref{thm:ppr_lb}, we rely heavily on the fact that when $\epsilon^2$ is sufficiently greater than $\frac{\sigma}{\rho}$, the normalized cut of $\Cset_1$ will be much larger than that of $\mathcal{L}$. In this case, not merely PPR but any algorithm which approximates the minimum normalized cut is unlikely to recover $\Cset_1$. Local spectral clustering algorithms based on truncated random walks \textcolor{red}{Spielman and Teng}, global spectral clustering algorithms \textcolor{red}{Shi and Malik}, and $p$-Laplacian based spectral embeddings \textcolor{red}{Buhler and Hein} all have provable upper bounds on the normalized cut of cluster they output, and thus it stands to reason that all the aforementioned approaches should struggle to estimate $\Cset_1[\Xbf]$.

@@

To better digest the implications of Theorem~\ref{thm:ppr_lb}, we apply our upper bound to the density function $f$ given in \eqref{eqn:lb_density}. Observe that $\Csig^{(1)}$ satisfies each of the assumptions~\textcolor{red}{(A1) - (A5).} In particular
\begin{enumerate}[label=(A\arabic*)]
	\item The density $f(x) = \frac{1 - \epsilon}{2 \rho \sigma}$ for all $x \in \Csig^{(1)}$.
	\item The clusters $\Csig^{(1)}$ and $\Csig^{(2)}$ are separated, $\dist(\Csig^{(1)}, \Csig^{(2)}) \geq \sigma$. 
	\item The density $f(x) = \frac{\epsilon}{\rho\sigma}$ for all $x$ such that $0 < \dist(x,\Csig^{(1)}) \leq \sigma$. Therefore for all such $x$,
	\begin{equation*}
	\inf_{x' \in \Csig^{(1)}} f(x') - f(x)  > \left\{\frac{1 - \epsilon}{2} - \epsilon \right\} \frac{1}{\rho \sigma}.
	\end{equation*}
	\item The set $\Csig^{(1)}$ is itself convex, and has diameter $\rho$.
	\item By symmetry, $\vol_{\Pbb,r}(\Csig^{(1)}) = \vol_{\Pbb,r}(\Csig^{(2)})$ and therefore $\vol_{\Pbb,r}(\Csig^{(1)}) \leq \frac{1}{2}\vol_{\Pbb,r}(\Reals^d)$. 
\end{enumerate}

%Technically, this density does not fit nicely into the setup of Theorem~\textcolor{red}{1}, because the sets $\Cset_{\sigma}^{(m)}$ ($ m = 1,2$) play the role of both the $\sigma$-expanded sets and the density clusters themselves. However, this truly is only a technical objection. One could modify the distribution $\Pbb$ by adding a small amount of mass to the $\sigma$-interior $\Cset^{(m)} = \set{x \in \Cset_{\sigma}^{(m)}: \dist(x,\partial\Cset_{\sigma}) \geq \sigma}$. Appropriately choosing a threshold $\lambda > \frac{(1 - \epsilon)}{2\rho\sigma}$, the resulting $\lambda$-density clusters would be $\Cbb_f(\lambda) = \set{\Cset^{(1)},\Cset^{(2)}}$, and their $\sigma$-expansions would resemble the rectangles $\Csig^{(1)}, \Csig^{(2)}$ except with slightly rounded corners. 
%Alternatively, one could directly modify Lemma~\textcolor{red}{6} -- an auxiliary Lemma, used in the proof of Theorem~\textcolor{red}{3}, which establishes an upper bound on the Lebesgue measure of $\Csig + B(0,r)$ -- to hold under the assumption $\Csig$ is a rectangle of width $\sigma$ rather than a $\sigma$-expansion. The proof of this Lemma is the only time we use the fact that $\Csig$ is a $\sigma$-expansion of a density cluster. \textcolor{red}{(Ryan and Siva.)}

\begin{remark}Technically, the rectangles $\Csig^{(1)}$ and $\Csig^{(2)}$ are not $\sigma$-expansions due to their sharp corners. To fix this, one can either modify the upper bound (specifically, Lemma~\textcolor{red}{6}) to hold with respect to rectangles of width $\sigma$, or change the sets $\Csig^{(1)}$ and $\Csig^{(2)}$ to be rectangles with rounded corners. As these details do not meaningfully change our conclusions, for simplicity we will keep our upper and lower bounds as is.
\end{remark}

If the user-specified parameters are initialized according to \textcolor{red}{(14)}, we may apply Theorem~\textcolor{red}{1}. This implies that there exists a set $\Csig^{(1)}[\Xbf] \subset \Csig^{(1)}$ with $\vol_{n,r}(\Csig[\Xbf]^g) \geq \frac{1}{2}\vol_{n,r}(\Csig[\Xbf])$ such that for any seed node $v \in \Csig^{(1)}[\Xbf]$, the estimated cluster $\widehat{C}$ output by Algorithm 1 satisfies the following upper bound:
\begin{equation*}
\vol_{n,r}(\widehat{C} \vartriangle \Csig^{(1)}[\Xbf]) \leq c \cdot \kappa(\Cset^{(1)}) \cdot \vol_{n,r}(\Csig^{(1)}[\Xbf])
\end{equation*}
where the condition number is given by
\begin{equation*}
\kappa(\Cset^{(1)}) = c \frac{\epsilon}{\sigma}\left(\frac{\rho^2}{r}\log^2\left(\frac{\Lambda_{\sigma}}{\lambda_{\sigma}^2r}\right) + c\right).
\end{equation*}

To facilitate comparisons between our upper and lower bounds, assume $\frac{1}{4}\sigma \leq \frac{1}{40}\rho$ and set $r = \frac{1}{4}\sigma$. 
For some universal constant $c > 0$, the following statements each hold with high probability:
\begin{itemize}
\item If the user-specified parameters satisfy \textcolor{red}{(14)} and for $b > 0$,
\begin{equation*}
\epsilon < c\frac{b\sigma^2}{\rho^2 \log^2\left(\frac{\Lambda_{\sigma}}{\lambda_{\sigma}^2\sigma}\right)}
\end{equation*}
then $\Delta(\widehat{C}, \mathcal{C}^{(1)}[\Xbf]) \leq b \cdot  \vol_{n,r}(\mathcal{C}^{(1)}[\Xbf])$.
\item If the user-specified parameters are given as in Theorem~\ref{thm:ppr_lb}, and 
\begin{equation*}
\epsilon > c\left({\frac{\sigma}{\rho}} \log\left(\frac{\rho \sigma}{\epsilon^2 r^2}\right)\right)^{1/4}
\end{equation*}
then $\Delta(\widehat{C}, \mathcal{C}^{(1)}[\Xbf]) \geq \frac{1}{8} \vol_{n,r}(\mathcal{C}^{(1)}[\Xbf])$.
\end{itemize}

Jointly, these upper and lower bounds give a relatively precise characterization of what it means for a density cluster to be well- or poorly-geometrically conditioned for recovery using PPR.

\begin{remark}
	It is worth pointing out that the above conclusions are reliant on specific (albeit reasonable) ranges and choices of input parameters, which in some instances differ between the upper and lower bounds. We suspect that our lower bound continues to hold even when choosing input parameters as dictated by our upper bound; however, proving this is currently beyond our technical grasp, and is a matter for future work. 
\end{remark}

%It is worth pointing out that both our lower and upper bounds are stated with respect to specific choices of the input parameters to Algorithm~\ref{alg:ppr}, and that these choices are in fact different in the case of the teleportation parameter $\alpha$ and sweep cut range $(L,U)$. At a high level, we do not believe this substantially weakens the takeaway messages of our work: if a density cluster $\mathcal{C}$ is geometrically well-conditioned, then a reasonable initialization of PPR will recover $\Csig[\Xbf]$ with low error, whereas if it is geometrically poorly-conditioned, then a different but also reasonable initialization of PPR will fail to recover $\Csig[\Xbf]$. Of course, it would be more satisfying if both our upper and lower bounds could be stated with respect to the same choices of these input parameters. Unfortunately, it is not at all obvious how to prove such a claim, as to the best of our knowledge all existing work on PPR assumes similar initialization conditions to the ones we assume; effectively, that the algorithm is well-initialized with respect to the set $S \subset G$ one is interested in recovering. We suspect that our lower bound continues to hold even when choosing input parameters as dictated by our upper bound; however, proving this is currently beyond our technical grasp, and is a matter for future work. Even so, we reiterate our belief that in tandem our upper and lower bounds represent a substantial step forward in characterizing the behavior of local clustering algorithms in a statistical context \textcolor{red}{(Ryan and Siva)}.

\begin{remark}
	We also note that our lower bound is of course specific to the PPR algorithm. As mentioned previously, density clustering is a well-studied problem, and in particular under our assumptions classical plug-in density cluster estimators can consistently recover the $\sigma$-expansion $\Csig$ of a density cluster $\Cset$, even when $\epsilon$ is large compared to $\frac{\sigma}{\rho}$. That PPR has difficulty recovering density clusters which can be easily estimated by standard plug-in approaches is not surprising, nor is it a knock on PPR. Rather, it simply reflects that while classical density clustering approaches are specifically designed to identify high-density regions regardless of their geometry, PPR considers geometry as well as density when deciding upon the optimal cluster.
\end{remark}

@@

To be precise, we have that there exists a set $\mathcal{L}[\Xbf]^g \subset \mathcal{L}[\Xbf]$ with $\vol_{n,r}(\mathcal{L}[\Xbf]^g) \geq \frac{5}{6} \vol_{n,r}(\mathcal{L}[\Xbf])$, such that the following statement holds for any $v \in \mathcal{L}[\Xbf]^g$: when Algorithm~\textcolor{red}{1} is run with inputs $\Xbf, r < \frac{1}{4}\sigma,\alpha = 65 \Phi_{\Pbb,r}(\mathcal{L}[\Xbf]),v \in \mathcal{L}[\Xbf]^g$ and $(L,U) = (0,1)$, the resulting PPR cluster estimate $\widehat{C}$ satisfies
\begin{align}
\Phi_{n,r}(\widehat{C}) & \leq \sqrt{11200\left\{\log\left(\frac{m}{d_{\min}^2}\right) + \log 20\right\} \Phi_{n,r}(\mathcal{L[X]})} \nonumber \\
& \leq \sqrt{89600\left\{\log\left(\frac{\rho \sigma}{\epsilon^2 \pi r^2}\right) + \log 20\right\} \frac{r}{\rho}} \label{eqn:lb_pf4} 
\end{align}
with probability at least $1 - 2n\exp\set{-\frac{\pi\epsilon r^2 n}{8978 \rho \sigma}} - 6 \exp\set{-\frac{1}{1123}(\cut_{\Pbb,r}(\mathcal{L}))^2n}$ (where the latter inequality follows from \eqref{eqn:lb_pf2} and Lemma~\ref{lem:graph_functional_concentration}.)

@@

The precise statement we will prove is contained in the following Lemma.
\begin{lemma}
	\label{lem:normalized_cut_lb}
	The normalized cut $\Phi_{n,r}(A)$ is upper bounded
	\begin{equation*}
	\Phi_{n,r}(A) \geq \frac{1}{12\pi} \left(1 - 4\frac{\sigma \rho}{r^2 n^2} \vol_{n,r}(A \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf]) \right) \frac{\epsilon^2 r}{\sigma}
	\end{equation*}
	uniformly over all $A \subset X$ with probability at least $1 - \exp\set{-2n\delta^2(\vol_{\Pbb,r}(\mathcal{X}))^2} - \frac{12\sigma \rho}{r^2} \exp\set{-\frac{\delta^2\epsilon r^2 n}{\rho \sigma(3 + \delta)}} - \frac{2\rho}{r}\exp\set{-\frac{\delta^2\pi r^3n}{2\sigma\rho^2(3 + \delta)}}$.
\end{lemma}
\begin{proof}
	To lower bound the normalized cut $\Phi_{n,r}(A)$, we must lower bound $\cut_{n,r}(A)$ and upper bound $\vol_{n,r}(A)$. A naive upper bound on the volume is simply 
	\begin{equation}
	\label{eqn:normalized_cut_lb_pf4}
	\vol_{n,r}(A) \leq \vol_{n,r}(G_{n,r}) \overset{(i)}{\leq} (1 + \delta)  \vol_{\Pbb,r}(\mathcal{X}) n^2 \leq (1 + \delta)\frac{\pi r^2}{\rho \sigma} n^2
 	\end{equation}
	where $(i)$ holds with probability at least $1 - \exp\set{-2n\delta^2(\vol_{\Pbb,r}(\mathcal{X}))^2}$, and it turns out this will suffice for our purposes. (Here and in the rest of this proof we take $\delta = 1/2$.)

@@

	It remains to lower bound the random quantities $R_{\min}$ and $Q_{\min}$.
	To do so, we first lower bound the expected probability of any cell $Q$,
	\begin{align*}
	\min_{Q \in \overline{Q}} \Pbb(Q) \geq \frac{\epsilon r^2}{\rho \sigma}.
	\end{align*}
	and the expected volume of $\mathcal{C}_{\sigma}^{(1)}[\Xbf] \cap R_k$ and $\mathcal{C}_{\sigma}^{(2)}[\Xbf] \cap R_k$,
	\begin{equation}
	\vol_{\Pbb,r}(\mathcal{C}_{\sigma}^{(1)} \cap R_k) = \vol_{\Pbb,r}(\mathcal{C}_{\sigma}^{(2)} \cap R_k) \geq \frac{\pi r^3}{2 \sigma \rho^2} \label{eqn:normalized_cut_lb_pf5}
	\end{equation}

	Since $Q_{\min}$ and $R_{\min}$ are obtained by taking the minimum of functionals over a fixed number of sets in $n$, they concentrate tightly around their means. Specifically, note that the total number of cubes is $\abs{\overline{Q}} = \frac{12 \sigma \rho}{r^2}$, and the total number of horizontal slices is $\frac{2\rho}{r}$. Along with \eqref{eqn:normalized_cut_lb_pf3} and \eqref{eqn:normalized_cut_lb_pf5}, by Lemma~\ref{lem:bernstein_union} 
	\begin{equation*}
	Q_{\min} \geq (1 - \delta)\frac{\epsilon r^2}{\rho \sigma} ~~\textrm{and}~~ R_{\min} \geq \frac{(1 - \delta)}{2}\frac{\pi r^3}{\sigma \rho^2 },
	\end{equation*}
	with probability at least $1 - \frac{12\sigma \rho}{r^2} \exp\set{-\frac{\delta^2\epsilon r^2 n}{\rho \sigma(3 + \delta)}} - \frac{2\rho}{r}\exp\set{-\frac{\delta^2\pi^2 r^6n}{4\sigma^2\rho^4}}$. Combining these lower bounds with \eqref{eqn:normalized_cut_lb_pf4} and \eqref{eqn:normalized_cut_lb_pf3}, we obtain
	\begin{equation*}	
	\Phi_{n,r}(A) \geq \frac{(1 - \delta)^2}{2(1 + \delta)\pi} \left(1 - 2 \frac{\sigma \rho}{(1 - \delta) r^2 n^2} \vol_{n,r}(A \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf]) \right) \frac{\epsilon^2 r}{\sigma},
	\end{equation*}
	and plugging in $\delta = 1/2$ yields the claim of Lemma~\ref{lem:normalized_cut_lb}.
\end{proof}

>>>>

As the following theorem demonstrates, even when Algorithm~\textcolor{red}{1} is reasonably initialized, if the density cluster $\mathcal{C}_{\sigma}^{(1)}$ is sufficiently geometrically ill-conditioned the cluster estimator $\widehat{C}$ will fail to recover $\mathcal{C}_{\sigma}^{(1)}$. Let
\begin{equation}
\label{eqn:lower_set}
\mathcal{L} = \set{(x_1,x_2) \in \mathcal{X}: x_2 < 0}.
\end{equation}
\begin{theorem}
	\label{thm:ppr_lb}
	Suppose $r < \frac{1}{40}\rho \wedge \frac{1}{4}\sigma$,  $\alpha = 65 \Phi_{\Pbb}(\mathcal{L})$, and $(L,U) = (0,1)$ are inputs to Algorithm~\textcolor{red}{1}. Then, for any
	\begin{equation}
	\label{eqn:lb_sample_size}
	n \geq \max\set{\frac{64}{\epsilon^2 \rho \sigma \pi r^2}, \frac{8}{\epsilon}}
	\end{equation}
	the following statement holds: there exists a set $C^{g} \subset \Xbf$ with $\vol(C^g \cap \mathcal{C}_{\sigma}^{(1)}[\Xbf]) \geq \frac{1}{4}\vol_{n,r}(\mathcal{C}_{\sigma}^{(1)}[\Xbf])$ such that for any seed node $v \in C^g$, the estimator $\widehat{C}$ computed by Algorithm~\textcolor{red}{1} has symmetric set difference with $\Csig^{(1)}[\Xbf]$ of volume at least
	\begin{equation}
	\label{eqn:ppr_lb}
	\frac{\sigma \rho}{r^2 n^2}\frac{\vol(\widehat{C} \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf])}{n} \geq \frac{1}{4} -  c \frac{\sqrt{\frac{\sigma}{\rho}}}{\epsilon^2} \sqrt{ \log\left(\frac{\rho \sigma}{\epsilon^2 r^2}\right)\frac{\sigma}{r}}
	\end{equation}
	with probability at least $1 - c_1 n \exp\set{-c_2n}$, where $c$ is a universal constant and $c_1,c_2$ are constants which do not depend on $n$.
\end{theorem}

Roughly speaking, our lower bound implies that if $\sigma\sqrt{\frac{1}{r\rho}} \gg \epsilon^2$, then PPR will fail to recover the empirical density cluster $\Cset_1[\Xbf]$, for many seed nodes $v \in \Cset_1[\Xbf]$.

Theorem~\ref{thm:ppr_lb} is stated with respect to a particular hard case, where the density clusters are rectangular subsets of $\Reals^2$. Although we picked this setting to make the Theorem statement simple, our results are easily generalized to $\Reals^d$ and to non-rectangular clusters. Additionally, although we state our lower bound with respect to PPR run on neighborhood graph, the conclusion is likely to hold for a much broader class of spectral clustering algorithms. In the proof of Theorem~\ref{thm:ppr_lb}, we rely heavily on the fact that when $\epsilon^2$ is sufficiently greater than $\frac{\sigma}{\rho}$, the normalized cut of $\Cset_1$ will be much larger than that of $\mathcal{L}$. In this case, not merely PPR but any algorithm which approximates the minimum normalized cut is unlikely to recover $\Cset_1$. Local spectral clustering algorithms based on truncated random walks \textcolor{red}{Spielman and Teng}, global spectral clustering algorithms \textcolor{red}{Shi and Malik}, and $p$-Laplacian based spectral embeddings \textcolor{red}{Buhler and Hein} all have provable upper bounds on the normalized cut of cluster they output, and thus it stands to reason that all the aforementioned approaches should struggle to estimate $\Cset_1[\Xbf]$.

@@

To better digest the implications of Theorem~\ref{thm:ppr_lb}, we apply our upper bound to the density function $f$ given in \eqref{eqn:lb_density}.

Technically, this density does not fit nicely into the setup of Theorem~\textcolor{red}{1}, because the sets $\Cset_{\sigma}^{(m)}$ ($ m = 1,2$) play the role of both the $\sigma$-expanded sets and the density clusters themselves. However, this truly is only a technical objection. One could modify the distribution $\Pbb$ by adding a small amount of mass to the $\sigma$-interior $\Cset^{(m)} = \set{x \in \Cset_{\sigma}^{(m)}: \dist(x,\partial\Cset_{\sigma}) \geq \sigma}$. Appropriately choosing a threshold $\lambda > \frac{(1 - \epsilon)}{2\rho\sigma}$, the resulting $\lambda$-density clusters would be $\Cbb_f(\lambda) = \set{\Cset^{(1)},\Cset^{(2)}}$, and their $\sigma$-expansions would resemble the rectangles $\Csig^{(1)}, \Csig^{(2)}$ except with slightly rounded corners. 
Alternatively, one could directly modify Lemma~\textcolor{red}{6} -- an auxiliary Lemma, used in the proof of Theorem~\textcolor{red}{3}, which establishes an upper bound on the Lebesgue measure of $\Csig + B(0,r)$ -- to hold with respect to rectangles of width $\sigma$ rather than $\sigma$-expansions. The proof of this Lemma is the only time we use the fact that $\Csig$ is a $\sigma$-expansion of a density cluster. \textcolor{red}{(Ryan and Siva.)}

Regardless, we will proceed to showing the behavior of our upper bound in this example. Observe that $\Csig^{(1)}$ satisfies each of the assumptions~\textcolor{red}{(A1) - (A5).} In particular
\begin{enumerate}[label=(A\arabic*)]
	\item The density $f(x) = \frac{1 - \epsilon}{2 \rho \sigma}$ for all $x \in \Csig^{(1)}$.
	\item The clusters $\Csig^{(1)}$ and $\Csig^{(2)}$ are separated, $\dist(\Csig^{(1)}, \Csig^{(2)}) \geq \sigma$. 
	\item The density $f(x) = \frac{\epsilon}{\rho\sigma}$ for all $x$ such that $0 < \dist(x,\Csig^{(1)}) \leq \sigma$. Therefore for all such $x$,
	\begin{equation*}
	\inf_{x' \in \Csig^{(1)}} f(x') - f(x)  > \left\{\frac{1 - \epsilon}{2} - \epsilon \right\} \frac{1}{\rho \sigma}.
	\end{equation*}
	\item The set $\Csig^{(1)}$ is itself convex, and has diameter $\rho$.
	\item By symmetry, $\vol_{\Pbb,r}(\Csig^{(1)}) = \vol_{\Pbb,r}(\Csig^{(2)})$ and therefore $\vol_{\Pbb,r}(\Csig^{(1)}) \leq \frac{1}{2}\vol_{\Pbb,r}(\Reals^d)$. 
\end{enumerate}
As long as the user-specified parameters are well-initialized, we may therefore apply Theorem~\textcolor{red}{1}. This implies that there exists a set $\Csig^{(1)}[\Xbf] \subset \Csig^{(1)}$ with $\vol_{n,r}(\Csig[\Xbf]^g) \geq \frac{1}{2}\vol_{n,r}(\Csig[\Xbf])$ such that for any seed node $v \in \Csig^{(1)}[\Xbf]$, the estimated cluster $\widehat{C}$ output by Algorithm 1 satisfies the following upper bound:
\begin{equation*}
\vol_{n,r}(\widehat{C} \vartriangle \Csig^{(1)}[\Xbf]) \leq c \cdot \kappa(\Cset^{(1)}) \cdot \vol_{n,r}(\Csig^{(1)}[\Xbf])
\end{equation*}
where the condition number is given by
\begin{equation*}
\kappa(\Cset^{(1)}) = c \frac{\epsilon}{\sigma}\left(\frac{\rho^2}{r}\log^2\left(\frac{\Lambda_{\sigma}}{\lambda_{\sigma}^2r}\right) + c\right).
\end{equation*}

Broadly speaking, our upper bound implies that when $\epsilon \ll \frac{\sigma r}{\rho^2}$, the volume of the symmetric set difference $\widehat{C} \vartriangle \Csig^{(1)}[\Xbf]$ will be small. To facilitate comparisons with our lower bound, assume $\frac{1}{4}\sigma \leq \frac{1}{40}\rho$ and set $r = \frac{1}{4}\sigma$. In this case, when $\epsilon \ll \frac{\sigma^2}{\rho^2}$ and the user-specified parameters satisfy \textcolor{red}{(14)}, then with high probability the resulting PPR cluster estimator $\widehat{C}$ will have small symmetric set difference with $\Csig^{(1)}[\Xbf]$. On the other hand, if $\epsilon \gg \left(\frac{\sigma}{\rho}\right)^{1/4}$ and the user-specified parameters satisfy the setup of Theorem~\ref{thm:ppr_lb}, then with high probability $\widehat{C}$ will have large symmetric set difference with $\Csig^{(1)}[\Xbf]$. Jointly, these upper and lower bounds give a relatively precise characterization of what it means for a density cluster to be well- or poorly-geometrically conditioned for recovery using PPR.

It is worth pointing out that both our lower and upper bounds are stated with respect to specific choices of the input parameters to Algorithm~\ref{alg:ppr}, and that these choices are in fact different in the case of the teleportation parameter $\alpha$ and sweep cut range $(L,U)$. At a high level, we do not believe this substantially weakens the takeaway messages of our work: if a density cluster $\mathcal{C}$ is geometrically well-conditioned, then a reasonable initialization of PPR will recover $\Csig[\Xbf]$ with low error, whereas if it is geometrically poorly-conditioned, then a different but also reasonable initialization of PPR will fail to recover $\Csig[\Xbf]$. Of course, it would be more satisfying if both our upper and lower bounds could be stated with respect to the same choices of these input parameters. Unfortunately, it is not at all obvious how to prove such a claim, as to the best of our knowledge all existing work on PPR assumes similar initialization conditions to the ones we assume; effectively, that the algorithm is well-initialized with respect to the set $S \subset G$ one is interested in recovering. We suspect that our lower bound continues to hold even when choosing input parameters as dictated by our upper bound; however, proving this is currently beyond our technical grasp, and is a matter for future work. Even so, we reiterate our belief that in tandem our upper and lower bounds represent a substantial step forward in characterizing the behavior of local clustering algorithms in a statistical context \textcolor{red}{Ryan and Siva}

\subsection{Sample complexity.}

\begin{itemize}
	\item We stress that our lower bound characterizes the difficulty PPR has in identifying geometrically poorly-conditioned density clusters, but not the overall hardness of the density clustering problem in a statistical sense. 
	\item It is a simple matter to prove that consistent estimation of the set $\Csig$ is possible. \textcolor{red}{(Ryan and Siva).}
	\item To the best of our knowledge minimax rates for density clustering are known only when the density function is Lipschitz smooth, whereas in our setup the density $f$ may not even be continuous. However, as a practical matter the extra cases included our setup are those where the clusters are very sharply defined, and therefore should be even easier to recover. See \textcolor{red}{Rinaldo and Wasserman} for a more thorough discussion on this matter.
	\item That PPR has difficulty recovering density clusters which can be easily estimated by standard plug-in approaches is not surprising, nor is it a knock on PPR. Rather, it simply reflects that while classical density clustering approaches are specifically designed to identify high-density regions regardless of their geometry, PPR considers geometry as well as density when deciding upon the optimal cluster.
\end{itemize}

@@

To be precise, we have that there exists a set $\mathcal{L}[\Xbf]^g \subset \mathcal{L}[\Xbf]$ with $\vol_{n,r}(\mathcal{L}[\Xbf]^g) \geq \frac{5}{6} \vol_{n,r}(\mathcal{L}[\Xbf])$, such that the following statement holds for any $v \in \mathcal{L}[\Xbf]^g$: when Algorithm~\textcolor{red}{1} is run with inputs $\Xbf, r < \frac{1}{4}\sigma,\alpha = 65 \Phi_{\Pbb,r}(\mathcal{L}[\Xbf]),v \in \mathcal{L}[\Xbf]^g$ and $(L,U) = (0,1)$, the resulting PPR cluster estimate $\widehat{C}$ satisfies
\begin{align}
\Phi_{n,r}(\widehat{C}) & \leq \sqrt{11200\left\{\log\left(\frac{m}{d_{\min}^2}\right) + \log 20\right\} \Phi_{n,r}(\mathcal{L[X]})} \nonumber \\
& \leq \sqrt{89600\left\{\log\left(\frac{\rho \sigma}{\epsilon^2 \pi r^2}\right) + \log 20\right\} \frac{r}{\rho}} \label{eqn:lb_pf4} 
\end{align}
with probability at least $1 - 2n\exp\set{-\frac{\pi\epsilon r^2 n}{8978 \rho \sigma}} - 6 \exp\set{-\frac{1}{1123}(\cut_{\Pbb,r}(\mathcal{L}[\Xbf]))^2n}$ (where the latter inequality follows from \eqref{eqn:lb_pf2} and Lemma~\ref{lem:graph_functional_concentration}.)

@@

The precise statement we will prove is contained in the following Lemma.
\begin{lemma}
	\label{lem:normalized_cut_lb}
	The normalized cut $\Phi_{n,r}(A)$ is upper bounded
	\begin{equation*}
	\Phi_{n,r}(A) \geq \frac{1}{12\pi} \left(1 - 4\frac{\sigma \rho}{r^2 n^2} \vol_{n,r}(A \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf]) \right) \frac{\epsilon^2 r}{\sigma}
	\end{equation*}
	uniformly over all $A \subset X$ with probability at least $1 - c_1\exp\{-c_2n\}$.
\end{lemma}
\begin{proof}
	To lower bound the normalized cut $\Phi_{n,r}(A)$, we must lower bound $\cut_{n,r}(A)$ and upper bound $\vol_{n,r}(A)$. A naive upper bound on the volume is simply 
	\begin{equation}
	\label{eqn:normalized_cut_lb_pf4}
	\vol_{n,r}(A) \leq \vol_{n,r}(G_{n,r}) \leq (1 + \delta)  \vol_{\Pbb,r}(\mathcal{X}) n^2 \leq (1 + \delta)\frac{\pi r^2}{\rho \sigma} n^2
 	\end{equation}
	where the last inequality holds with probability at least $1 - c_1\exp\{-c_2n\}$, and it turns out this will suffice for our purposes. (Here and in the rest of this proof we take $\delta = 1/2$.)

@@

	It remains to lower bound the random quantities $R_{\min}$ and $Q_{\min}$.
	To do so, we first lower bound the expected probability of any cell $Q$,
	\begin{align*}
	\min_{Q \in \overline{Q}} \Pbb(Q) \geq \frac{\epsilon r^2}{\rho \sigma}.
	\end{align*}
	and the expected volume of $\mathcal{C}_{\sigma}^{(1)}[\Xbf] \cap R_k$ and $\mathcal{C}_{\sigma}^{(2)}[\Xbf] \cap R_k$,
	\begin{equation*}
	\mathbb{E}(\vol_{n,r}\bigl(\mathcal{C}_{\sigma}^{(2)}[\Xbf] \cap R_k\bigr)) = \mathbb{E}(\vol_{n,r}\bigl(\mathcal{C}_{\sigma}^{(1)}[\Xbf] \cap R_k\bigr)) = \vol_{\Pbb,r}(\mathcal{C}_{\sigma}^{(1)} \cap R_k) \geq \frac{1}{2} \frac{\pi r^3}{\sigma \rho^2}
	\end{equation*}

	Since $Q_{\min}$ and $R_{\min}$ are obtained by taking the minimum of functionals over a fixed number of sets in $n$, applying Bernstein's inequality and a union bound gives
	\begin{equation*}
	Q_{\min} \geq (1 - \delta)\frac{\epsilon r^2}{\rho \sigma} ~~\textrm{and}~~ R_{\min} \geq \frac{(1 - \delta)}{2}\frac{\pi r^3}{\sigma \rho^2 },
	\end{equation*}
	with probability at least $1 - c_1 \exp\{-c_2 n\}$. Combining these lower bounds with \eqref{eqn:normalized_cut_lb_pf4} and \eqref{eqn:normalized_cut_lb_pf3}, we obtain
	\begin{equation*}	
	\Phi_{n,r}(A) \geq \frac{(1 - \delta)^2}{2(1 + \delta)\pi} \left(1 - 2 \frac{\sigma \rho}{(1 - \delta) r^2 n^2} \vol_{n,r}(A \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf]) \right) \frac{\epsilon^2 r}{\sigma}
	\end{equation*}
	with probability at least $1 - c_1 \exp\{-c_2 n\}$.
\end{proof}
